Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-12 examples.tex
+

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred,\rtv)  \ast  \texttt{r14}\mapsto_{r} \_ \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \ast \ulcorner \texttt{aligned va} \land  \theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred),\rtv \ast  \texttt{r14}\mapsto_{r} \_ \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \; \ast }_{\rtv}$
$\specline{\exists (\entryf ,\;\entrytr,\; \entrytw,\; \entryo: \texttt{word 64}) \ldotp \ulcorner (\texttt{addr\_L1 }\vaddr) = \entryo\urcorner \; \ast }_{\rtv}$

@@

$\specline{  \nfpointsto{\mask\vaddr\maskthree\entryf}{\mask\vaddr\maskthreeoff\entryf}\entrytr\qtwo\naddr \ast \nfpointsto{\mask\vaddr\masktwo\entrytr}{\mask\vaddr\masktwooff\entrytr}\entrytw\qthree\naddr \;\ast}_{\rtv}$
$\specline{\texttt{pte\_addr} \mapsto_{\texttt{vpte},\rtv} \entryo \;(\texttt{wzero 64}) \ast \texttt{rax}\mapsto_{r} \texttt{pte\_addr}  }_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred,\rtv) \ast  \texttt{r14}\mapsto_{r} \texttt{pte\_addr} \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \; \ast }_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred,\rtv) \ast  \texttt{r14}\mapsto_{r} \texttt{pte\_addr} \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \;\ast}_{\rtv}$

@@

$\specline{  \nfpointsto{\mask\vaddr\maskthree\entryf}{\mask\vaddr\maskthreeoff\entryf}\entrytr\qtwo\naddr \ast \nfpointsto{\mask\vaddr\masktwo\entrytr}{\mask\vaddr\masktwooff\entrytr}\entrytw\qthree\naddr \;\ast}_{\rtv}$
$\specline{\texttt{pte\_addr} \mapsto_{\texttt{vpte},\rtv} \entryo \;(\texttt{wzero 64}) \ast \texttt{rax}\mapsto_{r} \texttt{pte\_addr} \; \ast }_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred,\rtv) \ast  \texttt{r14}\mapsto_{r} \texttt{pte\_addr} \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \;\ast}_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta<[\vaddr:=\texttt{fpaddr}]>),\gammaPred,\rtv) \ast  \texttt{r14}\mapsto_{r} \texttt{pte\_addr} \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \; \ast}_{\rtv}$

@@

In Figure \ref{fig:mapping_code}, we see
%\todo[inline]{emphasize the write to the L1 page table is validated by the fact that the identity mapping exists}

\subsection{Unmapping a Page}
The reverse operation, unmapping a designated page that is currently mapped, is pleasingly the reverse: by flipping the entry back to an invalid entry, full ownership of the updated entry reverts back to the now-invalid cell of the L1 table.

\subsection{Change of Address Space}
A critical piece of \emph{trusted} code in verified OS kernels is the assembly code to change the current address space; current verified OS kernels currently lack effective ways to specify and reason about this low-level operation, for resaons outlined in Section \ref{sec:relwork}.

@@

@@

$\specline{\texttt{rdi+0} \mapsto_{v,\rtv} \_ \ast \texttt{rdi+8} \mapsto_{v,\rtv} \_ \ast \texttt{rdi+16} \mapsto_{v,\rtv} \_ \ast \texttt{rdi+24} \mapsto_{v,\rtv} \_ \ast \texttt{rdi+32} \mapsto_{v,\rtv} \_}_{\rtv}$
$\specline{\texttt{rdi+40} \mapsto_{v,\rtv} \_ \ast \texttt{rdi+48} \mapsto_{v,\rtv} \_\ast \texttt{rdi+56} \mapsto_{v,\rtv} \_}_{\rtv}$

@@

$\specline{\texttt{rdi+0} \mapsto_{v,\rtv} \texttt{rbxv} \ast \texttt{rdi+8} \mapsto_{v,\rtv} \texttt{rspv} \ast \texttt{rdi+16} \mapsto_{v,\rtv} \texttt{rbpv} \ast \texttt{rdi+24} \mapsto_{v,\rtv} \texttt{r12v} \ast}_{\rtv}$
$\specline{ \texttt{rdi+32} \mapsto_{v,\rtv} \texttt{r13v} \ast \texttt{rdi+40} \mapsto_{v,\rtv} \texttt{r14v} \ast \texttt{rdi+48} \mapsto_{v,\rtv} \texttt{r15v}\ast \texttt{rdi+56} \mapsto_{v,\rtv} \_}_{\rtv}$

@@

$\specline{\texttt{rdi+0} \mapsto_{v,\rtv} \texttt{rbxv} \ast \texttt{rdi+8} \mapsto_{v,\rtv} \texttt{rspv} \ast \texttt{rdi+16} \mapsto_{v,\rtv} \texttt{rbpv} \ast \texttt{rdi+24} \mapsto_{v,\rtv} \texttt{r12v} \ast }_{\rtv}$
$\specline{\texttt{rdi+32} \mapsto_{v,\rtv} \texttt{r13v} \ast \texttt{rdi+40} \mapsto_{v,\rtv} \texttt{r14v} \ast \texttt{rdi+48} \mapsto_{v,\rtv} \texttt{r15v}\ast \texttt{rdi+56} \mapsto_{v,\rtv} \rtv}_{\rtv}$    

>>>>

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred)  \ast  \texttt{r14}\mapsto_{r} \_ \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \ast \ulcorner \texttt{aligned va} \land  \theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred) \ast  \texttt{r14}\mapsto_{r} \_ \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \textsf{ pte\_addr} \; \ast }_{\rtv}$
$\specline{\exists (\entryf ,\;\entrytr,\; \entrytw,\; \entryo,\;\textsf{pte\_addr },\paddr \; : \texttt{word 64}) \ldotp \ulcorner \texttt{addr\_L1 }(\vaddr, \entryo) = \paddr \urcorner \; \ast }_{\rtv}$

@@

$\specline{  \nfpointsto{\mask\vaddr\maskthree\entryf}{\mask\vaddr\maskthreeoff\entryf}\entrytr\qtwo\naddr \ast \nfpointsto{\mask\vaddr\masktwo\entrytr}{\mask\vaddr\masktwooff\entrytr}\entryo\qthree\naddr \;\ast}_{\rtv}$
$\specline{\texttt{pte\_addr} \mapsto_{\texttt{vpte}} \paddr \;(\texttt{wzero 64}) \ast \texttt{rax}\mapsto_{r} \texttt{pte\_addr}  }_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred) \ast  \texttt{r14}\mapsto_{r} \texttt{pte\_addr} \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \textsf{ pte\_addr} \; \ast }_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred) \ast  \texttt{r14}\mapsto_{r} \texttt{pte\_addr} \ast \texttt{rdi}\mapsto_{r} \vaddr \;\ast}_{\rtv}$

@@

$\specline{  \nfpointsto{\mask\vaddr\maskthree\entryf}{\mask\vaddr\maskthreeoff\entryf}\entrytr\qtwo\naddr \ast \nfpointsto{\mask\vaddr\masktwo\entrytr}{\mask\vaddr\masktwooff\entrytr}\entryo\qthree\naddr \;\ast}_{\rtv}$
$\specline{\texttt{pte\_addr} \mapsto_{\texttt{vpte}} \paddr \;(\texttt{wzero 64}) \; \ast }_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred) \ast  \texttt{r14}\mapsto_{r} \texttt{pte\_addr} \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \;\ast}_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\<[\vaddr:=\texttt{fpaddr}]>\theta),\gammaPred) \ast  \texttt{r14}\mapsto_{r} \texttt{pte\_addr} \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{r} \_ \; \ast}_{\rtv}$

@@

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh page backing this virtual first requires an assurance on the existence L1 table entry which is realized in two folds:
\begin{itemize}
\item physical pointsto assertions on the tables L4, L3, and L2 reaching to the L1 level entry (l1e) ensures the successful table lookups (Specification Lines 4-6)
\item ,and a virtual L1 pointsto ($\mapsto_{\textsf{vpte}}$) on the address computed from the reached L1 entry ($ \ulcorner \texttt{addr\_L1}(\vaddr,\entryo) = \paddr\urcorner$) which is almost exactly same with the virtual-pointsto definition in Figure \ref{fig:virtualpointstosharing} except the physical page address is already computed, not exsitentially quantified:
 \[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{vpte}} \;\{\textsf{q}\} \; \paddr \; \vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \lambda \mathit{cr3val}\ldotp
    \exists \delta\ldotp
    \sumapaces{\mathit{cr3val}}\delta \ast 
  \sumwalkabs\vaddr\qfrac\paddr \ast \paddr \mapsto_{\mathsf{p}} \vpage
\end{array}
\]
\end{itemize}
After obtaining an virtual address \textsf{pte\_addr} in \textsf{rax} referring to a physical address, computed from already known L4-L2 physical memory pointstos ($\texttt{addr\_L1}(\vaddr,\entryo) = \paddr$),
in Specification Line 7 ($\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\;(\texttt{wzero 64})$), we save it to \textsf{r14} to be updated later in Line 9.

Then we allocate a fresh page initialized (i.e. zeroed) in Line 14 which return an \textsf{aligned} address of fresh page (\textsf{fpaddr}) in \textsf{rax} (Specification Line 19).

Therefore, when we unfold the $\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\;(\texttt{wzero 64})$ we will be able obtain a full ownerhsip on the L1 table physical pointsto $\paddr \mapsto_{\mathsf{p}} \textsf{wzero 64}$ which can be split to obtain a fractional ownership (q4) on it ($\paddr \mapsto_{\mathsf{p}} \{q4\} \;\textsf{wzero 64}$) to connect it with the L4-L2 physical table pointstos (Specification Lines 4-6) to obtain the complete physical table-walk assertion (L4\_L1\_PointsTo(rtv maddr l4e l3e l2e pa)).

After storing $\textsf{pte\_addr}$
%\todo[inline]{emphasize the write to the L1 page table is validated by the fact that the identity mapping exists}
\subsection{Unmapping a Page}
The reverse operation, unmapping a designated page that is currently mapped, is pleasingly the reverse: by flipping the entry back to an invalid entry, full ownership of the updated entry reverts back to the now-invalid cell of the L1 table.

\subsection{Change of Address Space}
A critical piece of \emph{trusted} code in verified OS kernels is the assembly code to change the current address space; current verified OS kernels currently lack effective ways to specify and reason about this low-level operation, for resaons outlined in Section \ref{sec:relwork}.

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\gammaPred) \ast [\rtv'+56](\mathcal{I}\texttt{ASpace}(\theta,\gammaPred') \ast \texttt{Pother})}_{\rtv}$

@@

$\specline{\texttt{rdi+0} \mapsto_{v} \_ \ast \texttt{rdi+8} \mapsto_{v} \_ \ast \texttt{rdi+16} \mapsto_{v} \_ \ast \texttt{rdi+24} \mapsto_{v} \_ \ast \texttt{rdi+32} \mapsto_{v} \_}_{\rtv}$
$\specline{\texttt{rdi+40} \mapsto_{v} \_ \ast \texttt{rdi+48} \mapsto_{v} \_\ast \texttt{rdi+56} \mapsto_{v} \_}_{\rtv}$

@@

$\specline{\texttt{rdi+0} \mapsto_{v} \texttt{rbxv} \ast \texttt{rdi+8} \mapsto_{v} \texttt{rspv} \ast \texttt{rdi+16} \mapsto_{v} \texttt{rbpv} \ast \texttt{rdi+24} \mapsto_{v} \texttt{r12v} \ast}_{\rtv}$
$\specline{ \texttt{rdi+32} \mapsto_{v} \texttt{r13v} \ast \texttt{rdi+40} \mapsto_{v} \texttt{r14v} \ast \texttt{rdi+48} \mapsto_{v} \texttt{r15v}\ast \texttt{rdi+56} \mapsto_{v} \_}_{\rtv}$

@@

$\specline{\texttt{rdi+0} \mapsto_{v} \texttt{rbxv} \ast \texttt{rdi+8} \mapsto_{v} \texttt{rspv} \ast \texttt{rdi+16} \mapsto_{v} \texttt{rbpv} \ast \texttt{rdi+24} \mapsto_{v} \texttt{r12v} \ast }_{\rtv}$
$\specline{\texttt{rdi+32} \mapsto_{v} \texttt{r13v} \ast \texttt{rdi+40} \mapsto_{v} \texttt{r14v} \ast \texttt{rdi+48} \mapsto_{v} \texttt{r15v}\ast \texttt{rdi+56} \mapsto_{v} \rtv}_{\rtv}$    
